The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-09-29 thesis/results.tex
thesis edits

As discussed in Chapters \ref{sec:introduction}, \ref{sec:observations},
we test whether Galactic tides can form the extended density profiles of
Scl and UMi. In this Chapter, we analyze tailored N-body simulations,
using the methods described in Chapter \ref{sec:methods}, to assess
tidal impacts on each galaxy. We find that tides drive dark matter loss
in both systems but leave their compact stellar components largely
unaffected. The Large Magellanic Cloud (LMC) perturbs the orbit of Scl,
and somewhat UMi, yet the resulting tidal effects are weaker than in a
Milky Way-only potential. Our simulations demonstrate that recent tides
are unlikely to alter the stellar structure of Scl or UMi.

@@

\subsection{Dark matter evolution}\label{dark-matter-evolution}

As a representation of the most extreme tidal history, we initially
investigate the \smallperi{} Scl model.

Scl experiences moderate tidal mass loss. Fig.~\ref{fig:scl_sim_images}
shows the stripping of dark matter and formation of diffuse streams
trailing and leading Sculptor's orbit. Because tidal stripping can be
described as removal of the lowest energy particles, most mass loss
occurs in the outer halo.

N-body models may deviate from a point-particle trajectory due to
dynamical self-friction \citep[e.g.,][]{miller+2020}. However, this
effect is slight for Scl, which ends near the observed position, without
adjusting the initial conditions (the green point in
Fig.~\ref{fig:scl_sim_images}).

The inner density cusp is tidally resilient.
Fig.~\ref{fig:scl_tidal_track} shows the initial and final circular
velocity profiles, and the evolution of the maximum circular velocity.
The maximum velocity drops from \(31\,\kms\) to \(22\,\kms\), evolving
along the tidal track from \citet{EN2021}. The final circular velocity
profile resembles the initial with an inner cusp, but has a sharper
outer truncation. Quantitatively, the halo loses \(>90\%\) its initial
mass, as listed in Table~\ref{tbl:scl_sim_results}. However, the inner
structure is not expected to be affected, as the Jacobi radius is over
3kpc, outside of the initial and final \(\rmax\). Thus, tides may remove
significant mass, but mostly from the outer halo.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.png}}
\caption[Sculptor tidal tracks]{Dark matter evolution for the
\smallperi{} model of Sculptor. Blue solid and orange dashed lines show
the initial and final circular velocity profiles. The points represent
the evolution of the maximum circular velocity, and the dotted black
line shows the tidal track from \citet{EN2021}. To calculate the
velocity profiles, unbound particles are iteratively removed,
recalculating the potential at each step assuming spherical
symmetry.}\label{fig:scl_tidal_track}
\end{figure}

@@

\subsection{Stellar evolution}\label{stellar-evolution}

Tides minimally affect the stellar component of Sculptor in the
\smallperi{} orbit. In Fig.~\ref{fig:scl_smallperi_i_f}, the projected
stellar distribution displays no prominent distortions, and the radial
density profile is nearly unchanged. Only at a surface density
\(\sim10^8\) times fainter than the centre may faint tidal features
emerge. The total stellar mass lost corresponds to \(\sim 10\) stars in
total (see Table~\ref{tbl:scl_sim_results})---a formidable challenge
with the best of observations.

More extended stellar profiles are more sensitive to tides. With Plummer
initial conditions, as shown in
Fig.~\ref{fig:scl_smallperi_plummer_i_f}, the model loses more stellar
mass and forms more luminous tidal tails. Observations reaching surface
densities \(\sim10\) times fainter than our data could reveal a stream
in this case. Nevertheless, the radial extent probed by our data remains
nearly unchanged by tidal evolution.

@@

We use the \texttt{L3M11} model of the MW and LMC potential from
\citet{vasiliev2024}. The \texttt{L3M11} potential is an evolving
multipole approximation of an N-body simulation including a live MW and
LMC dark matter halo. The potential includes a static MW bulge and disk,
evolving MW and LMC halos, and the MW reflex motion. In their
simulation, the MW was initially a NFW halo with \(r_s=16.5\,\)kpc and
\(M_{\rm 200}= 98.4\times10^{10}\Mo\) , and the LMC a NFW halo with
\(r_s=11.7\) and \(M_{200} = 24.6 \times 10^{10} \Mo\). The total
\texttt{L3M11} MW mass is lighter than our initial \citet{EP2020}
potential. Notably, this model has a previous LMC pericentre
\(\sim6\,\Gyr\) ago.

@@

The timing of the LMC encounter implies a break radius consistent with
the observed density kink. We select the orbit with the
\(6\sigma\)-smallest LMC pericentre. The orbit is selected following the
procedure in Section~\ref{sec:orbital_estimation} except with
observational uncertainties doubled. Results are similar when selecting
for a small MW pericentre instead. This \texttt{LMC-flyby} orbit, is
integrated back in time \(2\,\Gyr\) ago to isolate recent tidal effects.
We modify Scl's initial halo to have \(\rmax = 2.5\,\kpc\) and
\(\vmax = 25\,\kms\), slightly reducing the initial stellar velocity
dispersion. Fig.~\ref{fig:scl_lmc_orbits_effect} shows this selected
orbit in black and Table~\ref{tbl:orbit_ics} records the initial
conditions.

@@

The combined tidal force of the MW and LMC are even weaker for Scl than
in the MW-only case. Fig.~\ref{fig:scl_lmc_sim_images} shows the dark
matter evolution of Scl and the passage of the LMC. With only one MW
pericentre, Scl's dark matter is less disrupted than the previous
MW-only model. The subsequent LMC passage has little effect. As a
result, the dark matter structure evolves mildly and \(\sim 50\%\) of
mass remains bound (Table~\ref{tbl:scl_lmc_sims}).

Correspondingly, the stellar component is nearly unchanged.
Fig.~\ref{fig:scl_lmc_i_f} shows the projected stellar distributions and
density profiles of this model. While the break radius is within the
observed density profile, tidal effects are too weak to be detectible.
Structural properties of the stars similarly evolve little
(Table~\ref{tbl:scl_lmc_sim_stars}). While the instantaneous tidal force
from the LMC is larger than the MW, Scl does not experience the LMC
tidal field long enough to display disturbances.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_i_f.pdf}}
\caption[Sculptor initial and final density with LMC]{Similar to
Fig.~\ref{fig:scl_smallperi_i_f} except for the \texttt{LMC-flyby}
model. With only one MW pericentre and a recent, rapid LMC encounter,
tidal forces do not appear to affect the stellar
distribution.}\label{fig:scl_lmc_i_f}
\end{figure}

@@

We find, including only the MW, that tides only remove dark matter from
the outskirts of Scl. The central cusp and compact stellar distribution
are resilient to tides and any tidal effects are well-outside the reach
of current observations. We then find that the LMC strongly perturbs
Scl's orbit---in this case, Scl may be on first infall. However, with
only 2 pericentres, the combined tides of the LMC and MW tides are
weaker than for our initial model. In either case, tides are unlikely to
affect Sculptor's stellar component.

@@

The tidal evolution of Ursa Minor is similar to Sculptor in the MW-only
case. Fig.~\ref{fig:umi_sim_images} shows snapshots of the DM evolution.
UMi loses more DM mass, with only about 3\% of the total mass remaining
(see Fig.~\ref{fig:umi_tidal_track} and
Table~\ref{tbl:umi_sim_results}). However, tidal features in the
observed stellar components are still extremely faint, appearing outside
100 arcminutes in Fig.~\ref{fig:umi_smallperi_i_f}. Tides only
moderately affect the observed size and velocity dispersion of Ursa
Minor (Table~\ref{tbl:umi_sim_stars_results}).

A stream may just be beyond observational limits if Ursa Minor's stars
were initially more extended. As seen in Fig.~\ref{fig:umi_plummer_i_f},
the tidal tails are more luminous than for Sculptor and would begin near
the outermost density measurement. The properties of the stream are
shown in Fig.~\ref{fig:umi_tidal_stream}. The stream stars become
apparent outside of about 2 degrees, or the Jacobi radius. As expected
for tidal tails, the stream shows gradients in each observable, with a
dispersion in each observable similar to that of the dwarf progenitor.

@@

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_images.png}}
\caption[Ursa Minor simulation snapshots]{Similar to
Fig.~\ref{fig:scl_sim_images} except for Ursa Minor on the \smallperi{}
orbit.}\label{fig:umi_sim_images}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_stream.pdf}}
\caption[Ursa Minor predicted stream]{The properties of the stream
around the UMi \smallperi{} orbit with Plummer stars. The panels are all
as a function of \(\xi'\), the distance along the stream as defined by
the current GSR proper motion vector. The top panels show the GSR proper
motions in RA and Dec, and the bottom two show the distance and GSR
radial velocities. To sample the stream, we randomly draw 100,000
samples from the snapshot based on the stellar weights. A detectible
gradient in \(\mu_{\alpha*}\) and LOS velocity should be detectible if
the stream is tracked across several
degrees.}\label{fig:umi_tidal_stream}
\end{figure}

@@

Fig.~\ref{fig:umi_orbits_lmc} shows the effects of including an LMC on
the orbit of Ursa Minor. Predominantly, the effect is to increase the
orbital period, apocentre and pericentre. However, the orbit remains in
a similar plane and with similar morphology. As UMi is on the opposite
side of the Galaxy of the LMC, this is not unexpected. The deviation
from the MW-only orbit is mostly due to the LMC-induced reflex motion of
the Milky Way.

@@

@@

While tides affect UMi more strongly than Scl, the tidal effects are
insufficient to reshape the observed stellar density profile. Faint
tidal tails may be observable with deeper data. Finally, including the
LMC in the potential further weakens the tides experience by UMi.

@@

As Section~\ref{sec:results} shows, tides marginally affected the
stellar components of Scl and UMi, even under extreme orbits. While
Section~\ref{sec:results} only presents select models, alternative
initial conditions do not affect our qualitative conclusions on tidal
effects.

>>>>

As discussed in Chapters \ref{sec:introduction}, \ref{sec:observations},
we test whether Galactic tides are responsible for the extended density
profiles of Scl and UMi. In this Chapter, we analyze tailored N-body
simulations, using the methods described in Chapter \ref{sec:methods},
to assess the tidal impact of the Galactic potential. To anticipate our
main conclusion, we find that tides drive dark matter loss in both
systems but leave their compact stellar components largely unaffected.
The Large Magellanic Cloud (LMC) perturbs the orbit of Scl, and somewhat
UMi, yet the resulting tidal effects are weaker than in a Milky Way-only
potential. Our simulations demonstrate that recent tides are unlikely to
have altered the stellar structure of Scl or UMi.

@@

\subsection{Evolution of Sculptor's dark matter
halo}\label{evolution-of-sculptors-dark-matter-halo}

As a representation of an extreme tidal history, we initially
investigate the \smallperi{} Scl model.

Scl experiences moderate tidal mass loss after 10 Gyrs of evolution.
Fig.~\ref{fig:scl_sim_images} shows the stripping of dark matter and
formation of diffuse streams trailing and leading Sculptor's orbit.
Because tidal stripping can be described as a gradual removal of the
least bound particles, most mass loss occurs in the outer halo.

N-body models may deviate from a point-particle trajectory due to
dynamical self-friction \citep[e.g.,][]{white1983, miller+2020}.
However, this effect is slight for Scl, which ends near the observed
position, without adjusting the initial conditions (the green point in
Fig.~\ref{fig:scl_sim_images}).

The inner density cusp is tidally resilient.
Fig.~\ref{fig:scl_tidal_track} shows the initial and final circular
velocity profiles, and the evolution of the maximum circular velocity.
The maximum velocity drops from \(31\,\kms\) to \(22\,\kms\), evolving
along the tidal track from \citet{EN2021}. The final circular velocity
profile resembles the initial with an inner cusp, but has a sharper
outer truncation. Quantitatively, the halo loses \(>90\%\) its initial
mass, as listed in Table~\ref{tbl:scl_sim_results}. However, the inner
structure is not expected to be affected, as the Jacobi radius is over
3kpc, outside of the initial and final \(\rmax\). Thus, tides may remove
significant amounts of mass, but mostly from the outer halo.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.png}}
\caption[Sculptor tidal tracks]{Dark matter evolution for the
\smallperi{} model of Sculptor. Blue solid and orange dashed lines show
the initial and final circular velocity profiles. The points represent
the evolution of the maximum circular velocity, and the dotted black
line shows the tidal track from \citet{EN2021}. To calculate the
velocity profiles, unbound particles are iteratively removed,
recalculating the potential at each step assuming spherical symmetry.
\textbf{mark jacobi radius and stellar contributions\ldots, maybe sigmav
/ r\_h}}\label{fig:scl_tidal_track}
\end{figure}

@@

\subsection{Evolution of Sculptor's
stars}\label{evolution-of-sculptors-stars}

Tides minimally affect the stellar component of Sculptor in the
\smallperi{} orbit. In Fig.~\ref{fig:scl_smallperi_i_f}, the projected
stellar distribution displays no prominent distortions, and the radial
density profile is nearly unchanged. Only at a surface density
\(\sim10^8\) times fainter than the centre do some faint tidal features
emerge. The total stellar mass lost corresponds to \(\sim 10\) stars in
total (see Table~\ref{tbl:scl_sim_results})---a formidable challenge
with the best of observations.

Sculptor may not have been born as an exponential galaxy, and more
extended stellar profiles would be more sensitive to tides. As an
approximation of the present-day extended density profile, we initialize
a model with Plummer stars instead. As shown in
Fig.~\ref{fig:scl_smallperi_plummer_i_f}, the Plummer model loses more
stellar mass and forms more luminous tidal tails. Observations reaching
surface densities \(\sim10\) times fainter than our data could reveal a
stream in this case. Nevertheless, the radial extent probed by our data
remains nearly unchanged by tidal evolution.

The Jacobi and break radii further support that tidal effects should not
be apparent in the observed stellar component. As calculated for this
model, (see
Tables~\ref{tbl:scl_sim_results}, \ref{tbl:scl_sim_stars_results}), the
break and Jacobi radii both fall outside of \(\sim 100\) arcminutes for
either stellar component. Indeed, the stellar component only begins to
deviate from an exponential profile around this break radius
(Figs.~\ref{fig:scl_smallperi_i_f}, \ref{fig:scl_smallperi_plummer_i_f}).
Given that no orbits of Scl in this potential produce significantly
smaller break or Jacobi radii, it is unlikely any other orbit would
produce a significant, observable density excess.

@@

We use the \texttt{L3M11} model of the MW and LMC potential from
\citet{vasiliev2024}. The \texttt{L3M11} potential is an evolving
multipole approximation of an N-body simulation including a live MW and
LMC dark matter halo. The potential includes a static MW bulge and disk,
evolving MW and LMC halos, and the MW reflex motion. In their
simulation, the MW was initially a NFW halo with \(r_s=16.5\,\)kpc and
\(M_{\rm 200}= 98.4\times10^{10}\Mo\) , and the LMC a NFW halo with
\(r_s=11.7\) and \(M_{200} = 24.6 \times 10^{10} \Mo\). The total
\texttt{L3M11} MW mass is lighter than our initial \citet{EP2020}
potential.

@@

The timing of the LMC encounter implies a break radius consistent with
the beginning of the observed density excess. We select the orbit with
the \(6\sigma\)-smallest LMC pericentre. The orbit is selected following
the procedure in Section~\ref{sec:orbital_estimation} except with
observational uncertainties doubled. Results are similar when selecting
for a small MW pericentre instead. This \texttt{LMC-flyby} orbit, is
integrated back in time \(2\,\Gyr\) ago to isolate recent tidal effects.
We modify Scl's initial halo to have \(\rmax = 2.5\,\kpc\) and
\(\vmax = 25\,\kms\), slightly reducing the initial stellar velocity
dispersion. Fig.~\ref{fig:scl_lmc_orbits_effect} shows this selected
orbit in black and Table~\ref{tbl:orbit_ics} records the initial
conditions.

@@

The combined tidal effect of the MW and LMC are even weaker for Scl than
in the MW-only case. Fig.~\ref{fig:scl_lmc_sim_images} shows the dark
matter evolution of Scl and the passage of the LMC. With only one MW
pericentre, Scl's dark matter is less disrupted than the previous
MW-only model. The subsequent LMC passage modifies Scl's orbit but has
otherwise little effect. As a result, the dark matter structure evolves
mildly and \(\sim 50\%\) of mass remains bound
(Table~\ref{tbl:scl_lmc_sims}).

Correspondingly, the stellar component is nearly unchanged.
Fig.~\ref{fig:scl_lmc_i_f} shows the projected stellar distributions and
density profiles of this model. While the break radius is within the
observed density profile, tidal effects are too weak to be detectable.
Structural properties of the stars similarly evolve little
(Table~\ref{tbl:scl_lmc_sim_stars}).

While the instantaneous tidal force from the LMC is larger than the MW,
Scl does not experience the LMC tidal field long enough to display
disturbances. Scl and the LMC have a similar orbital plane, but orbit in
opposite directions (Fig.~\ref{fig:scl_lmc_orbits_effect}). As a result,
the interaction between Scl and the LMC is brief as their relative
velocity is \(\sim 400\,\kms\) at closest approach.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_i_f.pdf}}
\caption[Sculptor initial and final density with LMC]{Similar to
Fig.~\ref{fig:scl_smallperi_i_f} except for the \texttt{LMC-flyby}
model. The Jacobi and break radii here are calculated with respect to
the LMC, the corresponding radii with respect to the MW are larger. With
only one MW pericentre and a recent, rapid LMC encounter, tidal forces
do not appear to affect the stellar distribution. \textbf{todo fix break
radius position, orbits as arrows\ldots{}}}\label{fig:scl_lmc_i_f}
\end{figure}

@@

We find, including only the MW potential, that tides only remove dark
matter from the outskirts of Scl. The central cusp and compact stellar
distribution are resilient to tides and any tidal effects are
well-outside the reach of current observations. We then find that the
LMC strongly perturbs Scl's orbit---in this case, Scl may be on first
infall. However, with only 1 pericentre each for the LMC and MW, the
combined tides are weaker than for our initial model. In either case, we
conclude that tides are unlikely to affect Sculptor's stellar component.

@@

\subsection{Evolution of Ursa Minor's dark matter
halo}\label{evolution-of-ursa-minors-dark-matter-halo}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_images.png}}
\caption[Ursa Minor simulation snapshots]{Similar to
Fig.~\ref{fig:scl_sim_images} except for Ursa Minor on the \smallperi{}
orbit. Dark matter evolution is more dramatic than for
Scl.}\label{fig:umi_sim_images}
\end{figure}

@@

The tidal evolution of Ursa Minor is similar to Sculptor in the MW-only
potential. Fig.~\ref{fig:umi_sim_images} shows snapshots of the DM
evolution. UMi loses significantly more DM mass than Scl, forming
substantial dark matter streams encircling the MW several times.

UMi only retains 3\% of its total mass after 9 Gyr
(Table~\ref{tbl:umi_sim_results}). As a result, the final dark matter
component is much smaller than the initial, but still evolves along the
predicted tidal track Fig.~\ref{fig:umi_tidal_track}. Despite the more
substantial tidal evolution, the Jacobi radius is still large, lying at
\(\sim 4\,\kpc\), well beyond the final \(\rmax\).

Because of UMi's mass loss, the orbit deviates substantially from a
point orbit. Through our orbit-adjustment procedure in
Section~\ref{sec:orbit_corrections}, we recover nearly exactly the
present-day position of Ursa Minor by changing the initial positions by
\(20\,\kpc\) and \(\sim 9\,\kms\). These adjustments do not
significantly affect the qualitative structure or pericentre of the
orbit.

\subsection{Evolution of Ursa Minor's
stars}\label{evolution-of-ursa-minors-stars}

@@

@@

Tidal features in UMi's stellar component are still extremely faint,
becoming apparent only outside 100 arcminutes in
Fig.~\ref{fig:umi_smallperi_i_f}. Tides also only moderately affect the
observed size and velocity dispersion of Ursa Minor
(Table~\ref{tbl:umi_sim_stars_results}). For exponential initial
conditions, tidal effects are unlikely to ever be observable in the near
future.

Similar to Scl, we consider a model where UMi's stars are initially a
more extended Plummer profile, resembling the present-day density
profile. The stellar evolution of the Plummer stellar component is
similar to the exponential stellar profile (\citet{ig:umi_plummer_i_f}).
However, because there are more loosely-bound stars, the Plummer model
looses nearly 4\% of its initial stellar mass to tides
(Table~\ref{tbl:umi_sim_stars_results}), and tidal features may be
detectible if we measure densities 2 orders of magnitude fainter than
our present data. We show the properties of a stream in the Appendix
(Fig.~\ref{fig:umi_tidal_stream}), but such a stream is unlikely to be
observable in the near future.

@@

@@

Fig.~\ref{fig:umi_orbits_lmc} shows the effects of including an LMC on
the orbit of Ursa Minor. Predominantly, the effect is to increase the
orbital period, apocentre and pericentre. However, the orbit remains in
a similar plane and with similar morphology. As UMi is on the opposite
side as the Galaxy of the LMC, this is not unexpected.

The deviation from the MW-only orbit is mostly due to the LMC-induced
reflex motion of the Milky Way. Because the MW centre, and our origin,
is accelerated towards the LMC, UMi experiences a fictitious
acceleration opposite to the acceleration experienced by the MW centre,
causing the orbit to increase in characteristic radius.

@@

While tides affect UMi more strongly than Scl, the tidal effects are
insufficient to reshape the observed stellar density profile. Faint
tidal tails may be observable with deeper data. Finally, including the
LMC in the potential further weakens the tides experienced by UMi.

@@

As Section~\ref{sec:results} shows, tides marginally affected the
stellar components of Scl and UMi, even under extreme orbits. While
Section~\ref{sec:results} only presents select models, alternative
initial conditions do not affect our qualitative conclusions on tidal
effects expected for Scl and UMi in the Galactic (and LMC) potential.
